Subgroup detection in linear growth curve models with GLMM trees

### 2023-06-26 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
added new growth curve plot
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3264 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: fba33cd90a3336a2e4a788163a1fffcde8ef6392

@@

lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "tree", type = "simple", gp = gpar(cex = .7))
icc <- as.data.frame(VarCorr(lmm_tree))

@@

\begin{subfigure}{0.8\textwidth}%
<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
plot(lmm_tree_r, which = "tree", type = "simple", gp = gpar(cex = .7))
icc_r <- as.data.frame(VarCorr(lmm_tree_r))
lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
@
\caption{GLMM tree resulting from use of random-effects initialization.}
\end{subfigure}%

\begin{subfigure}{0.8\textwidth}%
<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
plot(lmm_tree_c, which = "tree", type = "simple", gp = gpar(cex = .7))
icc_c <- as.data.frame(VarCorr(lmm_tree_c))
lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
@
\caption{GLMM tree resulting from use of clustered covariances.}
\end{subfigure}%

@@

>>>>

<<growthplot, echo=FALSE, results=hide>>=
growthplot <- function(object, cluster = NULL, ...) {
  if(is.null(cluster)) cluster <- object$tree$fitted[["(cluster)"]]
  if(is.null(cluster)) cluster <- Formula::model.part(Formula::Formula(object$formula),
    data = model.frame(object$lmer), lhs = 0, rhs = 2L)[[1L]]
  plot(object$tree, terminal_panel = node_growthplot, tp_args = list(cluster = cluster, ...))
}

node_growthplot <- function(mobobj, cluster = NULL,
  id = TRUE, pop = TRUE, bg = "white",
  curvecol = gray(0.2, alpha = 1/6), curvelwd = 2,
  linecol = 2, linelwd = 3,
  ylines = NULL, xlab = FALSE, ylab = FALSE,
  mainlab = NULL, gp = gpar(...), ...)
{
  ## obtain dependent variable
  mf <- model.frame(mobobj)
  y <- Formula::model.part(mobobj$info$Formula, mf, lhs = 1L, rhs = 0L)
  if(isTRUE(ylab)) ylab <- names(y)
  if(identical(ylab, FALSE)) ylab <- ""
  if(is.null(ylines)) ylines <- ifelse(identical(ylab, ""), 3, 5)
  y <- y[[1L]]

  ## obtain explanatory variables
  X <- Formula::model.part(mobobj$info$Formula, mf, lhs = 0L, rhs = 1L)
  X <- X[,NCOL(X),drop=FALSE]
  xlab <- if(!identical(xlab, FALSE)) { if(isTRUE(xlab)) colnames(X) else xlab[1L] } else ""

  ## axis scaling
  xscale <- range(X) + c(-0.1, 0.1) * diff(range(X))
  yscale <- range(y) + c(-0.1, 0.1) * diff(range(y))

  ## other information
  fitted <- mobobj$fitted[["(fitted)"]]
  cf <- coef(mobobj)

  rval <- function(node) {

    ## node index
    nid <- id_node(node)
    ix <- fitted %in% nodeids(mobobj, from = nid, terminal = TRUE)

    ## subsets
    y <- y[ix]
    x <- X[ix, 1L]
    cf <- cf[as.character(nid), ]

    ## set up top viewport
    top_vp <- viewport(layout = grid.layout(nrow = 2, ncol = 3,
		       widths = unit(c(ylines, 1, 1), c("lines", "null", "lines")),
                       heights = unit(c(1, 1), c("lines", "null"))),
		       width = unit(1, "npc"),
                       height = unit(1, "npc") - unit(2, "lines"),
		       name = paste("node_growthplot", nid, sep = ""),
                       gp = gp)
    pushViewport(top_vp)
    grid.rect(gp = gpar(fill = bg, col = 0))

    ## main title
    top <- viewport(layout.pos.col = 2, layout.pos.row = 1)
    pushViewport(top)

    if (is.null(mainlab)) { 
      mainlab <- if(id) {
    	function(id, nobs) sprintf("Node %s (n = %s)", id, nobs)
      } else {
    	function(id, nobs) sprintf("n = %s", nobs)
      }
    }
    if (is.function(mainlab)) {
      mainlab <- mainlab(nid, info_node(node)$nobs)
    }
    grid.text(mainlab)
    popViewport()

    plot <- viewport(layout.pos.col = 2, layout.pos.row = 2,
                     xscale = xscale, yscale = yscale,
                     name = paste0("node_growthplot", nid, "plot"),
                     clip = FALSE)
    pushViewport(plot)

    grid.xaxis(at = c(ceiling(xscale[1L] * 10), floor(xscale[2L] * 10))/10)
    grid.yaxis(at = c(ceiling(yscale[1L] * 10), floor(yscale[2L] * 10))/10)
    grid.rect(gp = gpar(fill = "transparent"))
    grid.clip()

    if(!is.null(cluster)) {
      cluster <- factor(cluster[ix])
      for(j in levels(cluster)) {
        jx <- which(cluster == j)
        grid.lines(unit(x[jx], "native"), unit(y[jx], "native"), gp = gpar(col = curvecol, lwd = curvelwd))
      }
    }
    grid.abline(unit(cf[1], "native"), unit(cf[2], "native"), gp = gpar(col = linecol, lwd = linelwd))

    if(ylab != "") grid.text(ylab, y = unit(0.5, "npc"), x = unit(-2.5, "lines"), rot = 90)
    if(xlab != "") grid.text(xlab, x = unit(0.5, "npc"), y = unit(-2, "lines"))                
    if(pop) popViewport(2L) else upViewport(2L)

  }

  return(rval)
}
class(node_growthplot) <- "grapcon_generator"
@

@@

lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
growthplot(lmm_tree)
icc <- as.data.frame(VarCorr(lmm_tree))

@@

\begin{subfigure}{0.8\textwidth}%
<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
growthplot(lmm_tree_r)
icc_r <- as.data.frame(VarCorr(lmm_tree_r))
lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
@
\caption{GLMM tree resulting from use of random-effects initialization.}
\end{subfigure}%

\begin{subfigure}{0.8\textwidth}%
<<fig=TRUE, height=4.5, width=8, echo=FALSE>>=
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
growthplot(lmm_tree_c)
icc_c <- as.data.frame(VarCorr(lmm_tree_c))
lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
@
\caption{GLMM tree resulting from use of clustered covariances.}
\end{subfigure}%

@@

\begin{figure}
\caption{GLMM tree for linear growth curves of science abilities among 250~children in 3rd, 5th, and 8th grade.}

<<fig=TRUE, height=9, width=15, echo=FALSE>>=
growthplot(lmm_tree_c)
@

\label{fig:LMM_tree_c}
\end{figure}
